Frequency of exposure mediates the facilitation effect of cognateness in bilingual word acquisition

### 2023-01-18 manuscript/manuscript.qmd
New draft version, extend discussion, update references, add appendix, change title, fix cross-references

---
title: "The role of cross-linguistic lexical similarity on bilingual word acquisition"
editor: source
author:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - ref: upf
    corresponding: true
  - name: Daniela S. Ávila-Varela
    affiliations:
      - ref: upf
  - name: Ignacio Castillejo
    affiliations:
      - ref: uam
  - name: Núria Sebastian-Galles
    affiliations:
      - ref: upf
affiliations:
  - id: upf
    name: Center for Brain and Cognition, Universitat Pompeu Fabra
  - id: uam
    name: Departamento de Psicología, Universidad Autónoma de Madrid

@@

thanks: "This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances."
keywords: lexical acquisition, vocabulary, bilingualism
toc: false
fig-dpi: 1000
csl: "apa7.csl"
bibliography: "references.bib"
format:
  pdf:
    pdf-engine: lualatex
    linestretch: 1.5
    number-sections: true
    whitespace: small
---

@@

The foundations of word learning are in place early in age. Before the end of their first year of life infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [e.g., @fenson2007macarthurbates]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. These early stages of lexical acquisition are characterised by substantial variation across children reflected, for instance, on the variability of the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability] or on the proportion of acquired words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure, @bates1994developmental]. Despite this variability, children's trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10 to 16 month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

To date, most studies on early word acquisition have relied exclusively on data from monolingual children while neglecting, to a considerable extent, that a substantial proportion of the world population acquires more than one language since early ages [@grosjean2021extent]. Previous findings pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals, who only learnt English. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar--or even larger--number of words than monolinguals, only when the languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

@@

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically distant languages, like Spanish and English (Romance and Germanic, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning a pair of languages that share a high proportion of cognates should benefit more often from this facilitation effect than those learning a pair of languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@@

Floccia et al. pointed to *parallel activation* as a phenomenon that might potentially underpin their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals--unfamiliar with the Catalan translations of the Spanish words they uttered--did not show this effect. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Several subsequent studies have also provided similar evidence in comprehension in children [e.g., @vonholzen2012language; @poulin-dubois2013lexical]. Parallel activation is therefore a plausible explanation for Floccia et al.'s results: cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. This account is is line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words, already acquired or not [e.g., @hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. Importantly, bilinguals seem to be more likely to acquire words for which their translation equivalent has been already learnt in the other language, suggesting that semantic similarity also facilitates word acquisition across languages [@bilson2015semantic].

If phonological similarity also plays a cross-language facilitation role during word acquisition, cognate translation equivalents--which share both semantic and phonological similarity--should be acquired, on average, earlier that non-cognate translation equivalents--which share semantic similarity, but not phonological similarity. Evidence supporting and earlier age of acquisition for cognates is, to date, scarce. @bosch2014first used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (remarkably larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp. 70):

@@

We argue that the facilitative effect of cognateness on word acquisition can be explained by at least two mechanisms, illustrated in @fig-hypotheses. On the one hand, and in line with @mitchell2022cognates, the effect of cognateness might take place during early stages of word acquisition, before stable representations of any of the two word-forms of the translation equivalent has been formed. In this scenario, the auditory recognition any of the two word-forms might lead to the activation of its (cognate) translation in the other language via their phonological similarity. Since that both word-forms are likely to be heard in the presence of the same referent, the association between one of the word-forms and its referent could be strengthened after the child is exposed to that same word-form, or to its (similar-sounding) phonologically translation.

For instance, in the presence of a tree, an English-Spanish bilingual child might hear *flower* in some occasions, and *flor* in some others. Given the phonological similarity between *flower* and *flor*, it is possible that both word-forms are activated in the child's lexicon every time any of them is encountered. This would lead to an increase in the association strength between each word form and their common referent every time one of them is encountered, leading to an earlier age of acquisition. The case of a non-cognate translation equivalent, such as *cow* and *vaca* would be different. Although both word-forms are also encountered in the presence of the same referent, their low phonological similarity would not guarantee such facilitation when the child hears either of them, thus leading to a later age of acquisition for both. This account departs from the premise that a word-form can potentially activate its translation equivalent via phonology before acquisition, that is, before the association between such word-form and its referent is consolidated. This would allow both word-forms to mutually boost their acquisition after only few exposures. In this scenario, the effect of cognateness on acquisition should taken place early in age, and should be symmetric for both languages: the facilitation should occur in both word-forms simultaneously.

@@

In line with this second hypothesis, the size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure). Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have, on average, learnt the English word-form of translation equivalents earlier than the word-form in the additional language.  Under the second hypothesis, the acquisition of English words by English-dominant participants would then benefit less frequently from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form.

@@

In this study, we investigated the role of cognateness on the acquisition of translation equivalents, and if such effect is conditional to the amount of time the child is exposed to the corresponding language of each member of the translation pair. Using an online vocabulary checklist--designed specifically for this study--we collected data from a sample of children aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We adopted an Bayesian item response theory (*IRT*) approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to the word-form's cognate status, and the participant's frequency of exposure to the word-form (a composite measure of lexical frequency and language degree of exposure), while adjusting for other indices of word difficulty---lexical frequency [@goodman2008does] and length (phonemes)---and participant skill---age [see @kachergis2022standard for a similar approach].

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire [@garcia-castro2022multilex]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formR platform [@arslan2020formr]. This questionnaire is structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. In case two translation equivalents were possible for a given word, both were included as separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single item (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*). We included items from a diverse sample of `r n_categories` semantic/functional categories.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (`r format(n_te, big.mark = ",")` translation equivalents; `r format(n_item_language["catalan"], big.mark = ",")` items in Catalan, `r format(n_item_language["spanish"], big.mark = ",")` items in Spanish), while others filled a shorter version (~400 translation equivalents, ~400 items in Catalan, ~400 items in Spanish). These last families were randomly allocated into one of four different subsets of the complete list of items. These lists were designed so that each contained a representative sub-sample of the items from the complete list. Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists--were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. For the analyses in the present study, we considered responses to words corresponding to nouns, verbs, and adjectives [@fourtassi2020growth]. We excluded multi-word items (e.g., *barrita de cereales* [cereal bar]) and items that included more than one word-form (e.g., *mono/mico*).

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between `r format(dates[1], "%dth %B, %Y")` and `r format(dates[2], "%dth %B, %Y")`. `r n_participants[1]` of them participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra), and were contacted by e-mail or phone if their child were aged between `r floor(min(participants$age))` and `r floor(max(participants$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants-lp for a summary of the distribution of participants across ages and degrees of exposure to Catalan). All families gave informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device in a browser within the two week following the invitation to participate. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a small present.

```{r}
#| label: tbl-participants-lp
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan. Information about DoE was provided by families before filling the questionnaire. A 100% indicated that the participant was exclusively exposed to Catalan, and never to Spanish. A 0% indicates that the participant was not exposed to Catalan ever, and rather most of the time to Spanish. A 50% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100%, 0%, and 50% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25%-wide bins, and ages (in months) binned into 4 month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
participants %>%
    filter(id %in% df$id) %>% 
    mutate_at(vars(starts_with("doe_")), function(x) {
        floor(x * 10) / 10
    }) %>%
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%", "25-50%", "50-75%", "75-100%"),
            include.lowest = TRUE
        ) %>%
            as.character()
    ) %>%
    count(age, doe_catalan) %>%
    arrange(desc(doe_catalan)) %>%
    pivot_wider(names_from = age, values_from = n) %>%
    gt() %>%
    summary_rows(
        columns = 2:7,
        fns = list("Total" = ~ sum(., na.rm = TRUE)),
        formatter = fmt_integer
    ) %>%
    sub_missing(everything(), everything(), missing_text = "--") %>%
    cols_label(doe_catalan = "DoE Catalan") %>%
    tab_spanner("Age (months)", 2:8) %>%
    tab_footnote(
        "This proportion is complementary to the degree of exposure to Spanish, with the exception of those participants who were also exposed to a third language up to 10% of the time",
        cells_column_labels(columns = doe_catalan)
    ) %>%
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan)))
```

>>>>

---
title: "Frequency of exposure mediates the facilitation effect of cognateness in bilingual word acquisition"
editor: source
author:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - ref: upf
    corresponding: true
  - name: Daniela S. Ávila-Varela
    affiliations:
      - ref: upf
  - name: Ignacio Castillejo
    affiliations:
      - ref: uam
  - name: Núria Sebastian-Galles
    affiliations:
      - ref: upf
affiliations:
  - id: upf
    name: Center for Brain and Cognition, Universitat Pompeu Fabra
  - id: uam
    name: Departamento de Psicología, Universidad Autónoma de Madrid

@@

thanks: "This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances."
keywords: lexical acquisition, vocabulary, bilingualism
toc: false
fig-dpi: 1000
csl: "apa7.csl"
bibliography: "references.bib"
format:
  pdf:
    pdf-engine: lualatex
    number-sections: true
    whitespace: small
    linestretch: 1.5
---

@@

The foundations of word learning are in place early in age. Before the end of their first year of life infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [e.g., @fenson2007macarthurbates]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. These early stages of lexical acquisition are characterised by substantial variation across children reflected, for instance, on the variability of the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability] or on the proportion of acquired words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure; @bates1994developmental]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10 to 16 month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Despite the recent increase in language representation in the literature, most studies on early word acquisition have relied exclusively on data from monolingual children while neglecting, to a considerable extent, that a substantial proportion of the world population acquires more than one language since early ages [@grosjean2021extent]. Previous findings pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals, who only learnt English. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals, only when the languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

@@

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close like Spanish and Catalan (both Romance languages) they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically distant languages, like Spanish and English (Romance and Germanic, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning a pair of languages that share a high proportion of cognates should benefit more often from this facilitation effect than those learning a pair of languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@@

Floccia et al. pointed to *parallel activation* as a phenomenon that might potentially underpin their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, did not show this effect. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Subsequent studies have also provided similar evidence in comprehension in children [e.g., @vonholzen2012language; @poulin-dubois2013lexical]. Parallel activation is therefore a plausible explanation for Floccia et al.'s results: cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. This account is is line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. Importantly, bilinguals seem to be more likely to acquire words for which their translation equivalent has been already learnt in the other language, suggesting that semantic similarity also facilitates word acquisition across languages [@bilson2015semantic].

If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity, but not phonological similarity. Evidence supporting and earlier age of acquisition for cognates is, to date, scarce. @bosch2014first used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (remarkably larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp. 70):

@@

We argue that the facilitative effect of cognateness on word acquisition can be explained by at least two mechanisms, illustrated in @fig-hypotheses. On the one hand, and in line with @mitchell2022cognates, the effect of cognateness might take place during early stages of word acquisition, before stable representations of any of the two word-forms of the translation equivalent has been formed. In this scenario, the auditory recognition of any of the two word-forms might lead to the activation of its (cognate) translation in the other language via their phonological similarity. Since that both word-forms are likely to be heard in the presence of the same referent, the association between one of the word-forms and its referent could be strengthened after the child is exposed to that same word-form, or to its (similar-sounding) phonologically translation.

For instance, in the presence of a tree, an English-Spanish bilingual child might hear *flower* in some occasions, and *flor* in some others. Given the phonological similarity between *flower* and *flor*, it is possible that both word-forms are activated in the child's lexicon every time any of them is encountered. This would lead to an increase in the association strength between each word-form and their common referent every time one of them is encountered, leading to an earlier age of acquisition. The case of a non-cognate translation equivalent, such as *cow* and *vaca* would be different. Although both word-forms are also encountered in the presence of the same referent, their low phonological similarity would not guarantee such facilitation when the child hears either of them, thus leading to a later age of acquisition for both. This account departs from the premise that a word-form can potentially activate its translation equivalent via phonology before acquisition, that is, before the association between such word-form and its referent is consolidated. This would allow both word-forms to mutually boost their acquisition after only few exposures. In this scenario, the effect of cognateness on acquisition should taken place early in age, and should be symmetric for both languages: the facilitation should occur in both word-forms simultaneously.

@@

In line with this second hypothesis, the size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure). Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have on average learnt the English word-form of translation equivalents earlier than the word-form in the additional language.  Under the second hypothesis, the acquisition of English words by English-dominant participants would then benefit less frequently from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form.

@@

In this study, we investigated the role of cognateness on the acquisition of translation equivalents, and if such effect is conditional to the amount of time the child is exposed to the corresponding language of each member of the translation pair. Using an online vocabulary checklist designed specifically for this study we collected data from a sample of children aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We adopted an Bayesian item response theory (*IRT*) approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to the word-form's cognate status, and the participant's frequency of exposure to the word-form (a composite measure of lexical frequency and language degree of exposure), while adjusting for other indices of word difficulty like lexical frequency [@goodman2008does] and length (phonemes), and participant skill like age [see @kachergis2022standard for a similar approach].

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formR platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. In case two translation equivalents were possible for a given word, both were included as separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged into a single item (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*). We included items from a diverse sample of `r n_categories` semantic/functional categories.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). In total, each response to the questionnaire provided data for 400 translation equivalents, that is, for 800 words: 400 in Catalan and 400 in Spanish. Participants were randomly allocated into one of four different subsets of the complete list of items. These lists were designed so that each contained a representative sub-sample of the items from the complete list. Semantic/functional categories with less than 16 items---thus resulting in less than four items after dividing it in four lists---were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. For the analyses in the present study, we considered responses nouns [@fourtassi2020growth]. We excluded multi-word items (e.g., *barrita de cereales* [cereal bar]) and items that included more than one word-form (e.g., *mono/mico*).

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between `r format(dates[1], "%dth %B, %Y")` and `r format(dates[2], "%dth %B, %Y")`. `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra), and were contacted by e-mail or phone if their child were aged between `r floor(min(df$age))` and `r ceiling(max(df$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan [see @tbl-participants for a summary of the distribution of participants across ages and degrees of exposure to Catalan]. All families gave informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device in a browser within the two week following the invitation to participate. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a small present.

```{r}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan. Information about DoE was provided by families before filling the questionnaire. A 100% indicated that the participant was exclusively exposed to Catalan, and never to Spanish. A 0% indicates that the participant was not exposed to Catalan ever, and rather most of the time to Spanish. A 50% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100%, 0%, and 50% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25%-wide bins, and ages (in months) binned into 4 month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
participants %>%
    filter(id %in% df$id) %>% 
    mutate_at(vars(starts_with("doe_")), function(x) {
        floor(x * 10) / 10
    }) %>%
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%", "25-50%", "50-75%", "75-100%"),
            include.lowest = TRUE
        ) %>%
            as.character()
    ) %>%
    count(age, doe_catalan) %>%
    arrange(desc(doe_catalan)) %>%
    pivot_wider(names_from = age, values_from = n) %>%
    gt() %>%
    summary_rows(
        columns = 2:7,
        fns = list("Total" = ~ sum(., na.rm = TRUE)),
        formatter = fmt_integer
    ) %>%
    sub_missing(everything(), everything(), missing_text = "--") %>%
    cols_label(doe_catalan = "DoE Catalan") %>%
    tab_spanner("Age (months)", 2:8) %>%
    tab_footnote(
        "This proportion is complementary to the degree of exposure to Spanish, with the exception of those participants who were also exposed to a third language up to 10% of the time",
        cells_column_labels(columns = doe_catalan)
    ) %>%
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan)))
```

We used the highest educational attainment of parents or caretakers as a proxy of participants' socio-economic status (SES), which families self-reported in the questionnaire by filling two items asking for the educational attainment of each parent or caretaker, with the following available options: *No education*, *Primary*, *Secondary*, *Complementary*, *Vocational*, and *University*, in line with the current educational system in Spain. Most families reported university studies (`r table(participants$edu_parent)["University"]`, `r percent(table(participants$edu_parent)["University"]/length(unique(participants$id)))`), followed by families were the highest educational attainment were vocational studies (`r table(participants$edu_parent)["Vocational"]`, `r percent(table(participants$edu_parent)["Vocational"]/length(unique(participants$id)))`), complementary studies (`r table(participants$edu_parent)["Complementary"]`,  `r percent(table(participants$edu_parent)["Complementary"]/length(unique(participants$id)))`), secondary education (`r table(participants$edu_parent)["Secondary"]`, `r percent(table(participants$edu_parent)["Secondary"]/length(unique(participants$id)))`), no formal education (`r table(participants$edu_parent)["No education"]`, `r percent(table(participants$edu_parent)["No education"]/length(unique(participants$id)))`) and primary education (`r table(participants$edu_parent)["Primary"]`, <1%).
